Storm surge modelling in the Strait of Georgia

### 2014-07-22 RoughWork/manuscript.tex
Worked on introduction

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are particularly susceptible to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. Storm surge hindcasts have been produced using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) in a regional configuration. The model has been evaluated through comparisons of tidal elevation harmonics, tidal currents, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly.
%needs a little more work.
\end{abstract}

@@

%my attempt at physical oceanopgraphy SoG- Susan feel free to edit.
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait. Several features related to the physical oceanopgraphy of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models

@@

Communities in the Strait of Georgia are partiuclarly susceptible to storm surges in the winter months when there is an increased propensity for storms and south easterly winds and geostrophic adjustment lead to higher coastal waters \citep{danard2003storm}. In the summer months, the wind direction changes to north westerly and mean sea level is lower. 
%work on this

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are particularly susceptible to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly.
%needs a little more work.
\end{abstract}

@@

%my attempt at physical oceanopgraphy in SoG- Susan feel free to edit.
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait. Several features related to the physical oceanopgraphy of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early efforts by \citet{crean1988numerical, stronach1993update} implemented a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. 
%tide models (Foreman), baroclinic (Masson, PNWL, Sutherland and MacCready)

@@

Communities in the Strait of Georgia are partiuclarly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal range is typically observed in the winter months during the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges are typically observed between Novemeber and Feburary. Several historical examples have been documented by \citet{forseth2006adaptation}. 
%work on this
